The Meta-Ontology of AI systems with Human-Level Intelligence

### 2023-02-06 ART_Krzanowski_Polak/Krzanowski_Polak.tex
Final commit before sending to the editors for corrections

\begin{artengenv2auth}{Roman Krzanowski, Pawel Polak}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
	%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
	\parencite[][p.57]{smith_promise_2019}. %
	 In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
	%\label{ref:RND1W3b57kSJx}(2019).
	\parencite*[][]{smith_promise_2019}.%
	}
	{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, J. Haugeland, B. C. Smith, M. Minsky, H. Dreyfus.}
	{%
		{\flushright\subbold{Roman Krzanowski}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
		{\flushright\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
	}

%\begin{document}
%\title{The meta-ontology of AI systems with human-level intelligence}
%\maketitle
%
%\section{Abstract}
%In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
%%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
%\parencite[][p.57]{smith_promise_2019}. %
% In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
%%\label{ref:RND1W3b57kSJx}(2019).
%\parencite*[][]{smith_promise_2019}.%
%
%
%Keywords: \textit{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, J. Haugeland, B. C. Smith, M. Minsky, H. Dreyfus.}

\section{Introduction}
AI systems have developed over the past 60 years, bringing new solutions to a~huge number of practical problems, and they continue to find many surprising and fascinating applications. However, the main goal of AI, namely the creation of a~human-like intelligence, \footnotetext{See ft. 3 on AGI for an explanation of human-like intelligence.} is still proving unattainable. Indeed, the
AI systems we currently design and implement cannot replicate human intelligence and a~human agent's ability to cope with reality
%\label{ref:RNDzbUfjwINVD}(see e.g. Brooks, 1991; Minsky, 1991; Dreyfus, 2016; Mitchell, 2019; Bołtuć, 2020; Roitblat, 2020; Wooldridge, 2021).
\parencites[see, e.g.,][]{brooks_intelligence_1991}[][]{minsky_logical_1991}[][]{dreyfus_skillful_2016}[][]{mitchell_artificial_2019}[][]{boltuc_conscious_2020}[][]{roitblat_algorithms_2020}[][]{wooldridge_road_2021}.%

@@

\section{Key grounding ideas}
\subsection{The ontology of AI}

Ontology can be thought as an empty buzzword or a~specific conceptual construct,\footnote{``We must be careful in reading [auth. any] philosophical works on ontology, when the author speaks of ‘ontology' without qualifications, not to confuse the intended sense of the world with any of the alternatives''
%\label{ref:RNDfLy5S3W893}(Jacquette, 2002, p.3).
\parencite[][p.3]{jacquette_ontology_2002}. %
 There is also a~confusion between ontology and metaphysics. Some authors see ontology as the ultimate study of reality 
%\label{ref:RNDvlH3CHHmmY}(e.g. Jacquette, 2002; Stróżewski, 2004, p.32, or Perzanowski, 2015),
\parencites[e.g.,][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[or][]{perzanowski_rozprawa_2015}, %
 and metaphysics as being ``after physics'', some others see ontology as a~part of metaphysics 
%\label{ref:RNDeDepxi7xRN}(see e.g. Van Inwagen, 2009).
\parencite[see e.g.][]{van_inwagen_metaphysics_2009}. %
 In AI literature because ontology takes on a~very concrete garb (of an engineering domain) metaphysics is a~rare term so the confusion is not so visible.} so we need to position the ontology of AI\textsubscript{E~}within the world of ontological theories and demonstrate, what it means in this discussion, how it relates to ``other'' ontologies in philosophy, computer science, and AI. Ontology in computer science and AI systems 
%\label{ref:RNDxG2UeatxFk}(e.g. Sánchez, Cavero and Martínez, 2007; as well Guarino and Giaretta, 1995; Guarino, Oberle and Staab, 2009; Swar, Khoriba and Belal, 2022)
\parencites[e.g.,][]{sharman_road_2007}[as well][]{guarino_ontologies_1995}[][]{guarino_what_2009}[][]{swar_unified_2022} %
 takes on different meanings to that of philosophy 
%\label{ref:RNDWOSOcJkOvq}(e.g. Jacquette, 2002; Stróżewski, 2004; Baker, 2007; Chalmers, Wasserman and Manley, 2009; Effingham, 2013; Berto and Plebani, 2015; Perzanowski, 2015; Thomasson, 2015; Hofweber, 2021).
\parencites[e.g.,][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[][]{baker_metaphysics_2007}[][]{chalmers_metametaphysics_2009}[][]{effingham_introduction_2013}[][]{berto_ontology_2015}[][]{perzanowski_rozprawa_2015}[][]{thomasson_ontology_2015}[][]{hofweber_logic_2021}.%
\footnote{Importing AI (or technical) aspects into philosophy brings with it a~touch of reality that philosophical considerations often lack. See also the comment by Jacquette on the relation between a~domain ontology and the domain itself 
%\label{ref:RNDY4CVeRGcfL}(Jacquette, 2002, p.5).
\parencite[][p.5]{jacquette_ontology_2002}.%
} The philosophical concepts of ontology, however, are fundamental to those used in specific applications 
%\label{ref:RNDFJ4JNj0TbG}(see the comments of Jacquette, 2002, p.XII).
\parencite[see the comments of][p.XII]{jacquette_ontology_2002}. %
 This is therefore where we begin.

@@

Ontology in philosophy may also refer to ``what exists'' in a~much more constrained, narrower, sense. This kind of ontology investigates existing, subject to a~definition for existence, objects and relations in the world, and we will refer to this ontology as O. Depending on the assumptions made, different types of objects and relations may be recognized by O~ontologies, because O~branches into many subdomains. Thus, many different perspectives have been developed for ontology
%\label{ref:RNDJ0HiQvGuma}(e.g. Quine, 1960; Jacquette, 2002; Stróżewski, 2004; Baker, 2007; Chalmers, Wasserman and Manley, 2009; Effingham, 2013; Ingarden, 2013; 2016; Berto and Plebani, 2015; Perzanowski, 2015; Thomasson, 2015; Hofweber, 2021);
\parencites*[e.g.,][]{quine_word_1960}[][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[][]{baker_metaphysics_2007}[][]{chalmers_metametaphysics_2009}[][]{effingham_introduction_2013}[][]{ingarden_controversy_2013}[][]{ingarden_controversy_2016}[][]{berto_ontology_2015}[][]{perzanowski_rozprawa_2015}[][]{thomasson_ontology_2015}[][]{hofweber_logic_2021}; %
 these differ in terms of extent, content, consistency, and accuracy, often responding to the specific needs of a~domain.

@@

Thus, ontology that represents the real world being always expressed in some form of language (which always is) is always incomplete with respect to the world, i.e., the world ``as is'' cannot be represented by something else, despite the close isomorphism between the reality and some logical aka ontological systems \footnote{All pure ontological systems are logical systems
%\label{ref:RNDQXsIfZlLYA}(Jacquette, 2002, p.xiii; see e.g. Foschini, 2013).
\parencites[][p.xiii]{jacquette_ontology_2002}[see, e.g.,][]{foschini_where_2013}.%
}

\end{artengenv2auth}

>>>>

\begin{artengenv2auth}{Roman Krzanowski, Pawel Polak}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{\label{krzanowpolak_start}In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
	%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
	\parencite[][p.57]{smith_promise_2019}. %
	 In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
	%\label{ref:RND1W3b57kSJx}(2019).
	\parencite*[][]{smith_promise_2019}.%
	}
	{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, J. Haugeland, B. C. Smith, M. Minsky, H. Dreyfus.}
	{%
%		{\flushright\subbold{Roman Krzanowski}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
%		{\flushright\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
		{\flushright\subbold{Roman Krzanowski}\\\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
	}

%\begin{document}
%\title{The meta-ontology of AI systems with human-level intelligence}
%\maketitle
%
%\section*{Abstract}
%In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
%%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
%\parencite[][p.57]{smith_promise_2019}. %
% In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
%%\label{ref:RND1W3b57kSJx}(2019).
%\parencite*[][]{smith_promise_2019}.%
%
%
%Keywords: \textit{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, J. Haugeland, B. C. Smith, M. Minsky, H. Dreyfus.}

\section*{Introduction}
\lettrine[loversize=0.13,lines=2,lraise=-0.07,nindent=5pt,findent=-5pt]%
{A}{}rtificial intelligence systems have developed over the past 60 years, bringing new solutions to a~huge number of practical problems, and they continue to find many surprising and fascinating applications. However, the main goal of AI, namely the creation of a~human-like intelligence,\footnote{See ft. 3 on AGI for an explanation of human-like intelligence.} is still proving unattainable. Indeed, the
AI systems we currently design and implement cannot replicate human intelligence and a~human agent's ability to cope with reality
%\label{ref:RNDzbUfjwINVD}(see e.g. Brooks, 1991; Minsky, 1991; Dreyfus, 2016; Mitchell, 2019; Bołtuć, 2020; Roitblat, 2020; Wooldridge, 2021).
\parencites[see, e.g.,][]{brooks_intelligence_1991}[][]{minsky_logical_1991}[][]{dreyfus_skillful_2016}[][]{mitchell_artificial_2019}[][]{boltuc_conscious_2020}[][]{roitblat_algorithms_2020}[][]{wooldridge_road_2021}.%

@@

\section*{Key grounding ideas}
\subsection*{The ontology of AI}

Ontology can be thought as an empty buzzword or a~specific conceptual construct,\footnote{``We must be careful in reading [auth. any] philosophical works on ontology, when the author speaks of ‘ontology' without qualifications, not to confuse the intended sense of the world with any of the alternatives''
%\label{ref:RNDfLy5S3W893}(Jacquette, 2002, p.3).
\parencite[][p.3]{jacquette_ontology_2002}. %
 There is also a~confusion between ontology and metaphysics. Some authors see ontology as the ultimate study of reality 
%\label{ref:RNDvlH3CHHmmY}(e.g. Jacquette, 2002; Stróżewski, 2004, p.32, or Perzanowski, 2015),
\parencites[e.g.,][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[or][]{perzanowski_rozprawa_2015}, %
 and metaphysics as being ``after physics'', some others see ontology as a~part of metaphysics 
%\label{ref:RNDeDepxi7xRN}(see e.g. Van Inwagen, 2009).
\parencite[see e.g.,][]{van_inwagen_metaphysics_2009}. %
 In AI literature because ontology takes on a~very concrete garb (of an engineering domain) metaphysics is a~rare term so the confusion is not so visible.} so we need to position the ontology of AI\textsubscript{E~}within the world of ontological theories and demonstrate, what it means in this discussion, how it relates to ``other'' ontologies in philosophy, computer science, and AI. Ontology in computer science and AI systems 
%\label{ref:RNDxG2UeatxFk}(e.g. Sánchez, Cavero and Martínez, 2007; as well Guarino and Giaretta, 1995; Guarino, Oberle and Staab, 2009; Swar, Khoriba and Belal, 2022)
\parencites[e.g.,][]{sharman_road_2007}[as well][]{guarino_ontologies_1995}[][]{guarino_what_2009}[][]{swar_unified_2022} %
 takes on different meanings to that of philosophy 
%\label{ref:RNDWOSOcJkOvq}(e.g. Jacquette, 2002; Stróżewski, 2004; Baker, 2007; Chalmers, Wasserman and Manley, 2009; Effingham, 2013; Berto and Plebani, 2015; Perzanowski, 2015; Thomasson, 2015; Hofweber, 2021).
\parencites[e.g.,][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[][]{baker_metaphysics_2007}[][]{chalmers_metametaphysics_2009}[][]{effingham_introduction_2013}[][]{berto_ontology_2015}[][]{perzanowski_rozprawa_2015}[][]{thomasson_ontology_2015}[][]{hofweber_logic_2021}.%
\footnote{Importing AI (or technical) aspects into philosophy brings with it a~touch of reality that philosophical considerations often lack. See also the comment by Jacquette on the relation between a~domain ontology and the domain itself 
%\label{ref:RNDY4CVeRGcfL}(Jacquette, 2002, p.5).
\parencite[][p.5]{jacquette_ontology_2002}.%
} The philosophical concepts of ontology, however, are fundamental to those used in specific applications 
%\label{ref:RNDFJ4JNj0TbG}(see the comments of Jacquette, 2002, p.XII).
\parencite[see the comments of][p.XII]{jacquette_ontology_2002}. %
 This is therefore where we begin.

@@

Ontology in philosophy may also refer to ``what exists'' in a~much more constrained, narrower, sense. This kind of ontology investigates existing, subject to a~definition for existence, objects and relations in the world, and we will refer to this ontology as O. Depending on the assumptions made, different types of objects and relations may be recognized by O~ontologies, because O~branches into many subdomains. Thus, many different perspectives have been developed for ontology
%\label{ref:RNDJ0HiQvGuma}(e.g. Quine, 1960; Jacquette, 2002; Stróżewski, 2004; Baker, 2007; Chalmers, Wasserman and Manley, 2009; Effingham, 2013; Ingarden, 2013; 2016; Berto and Plebani, 2015; Perzanowski, 2015; Thomasson, 2015; Hofweber, 2021);
\parencites[e.g.,][]{quine_word_1960}[][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[][]{baker_metaphysics_2007}[][]{chalmers_metametaphysics_2009}[][]{effingham_introduction_2013}[][]{ingarden_controversy_2013}[][]{ingarden_controversy_2016}[][]{berto_ontology_2015}[][]{perzanowski_rozprawa_2015}[][]{thomasson_ontology_2015}[][]{hofweber_logic_2021}; %
 these differ in terms of extent, content, consistency, and accuracy, often responding to the specific needs of a~domain.

@@

%Thus, ontology that represents the real world being always expressed in some form of language (which always is) is always incomplete with respect to the world, i.e., the world ``as is'' cannot be represented by something else, despite the close isomorphism between the reality and some logical aka ontological systems \footnote{All pure ontological systems are logical systems
%%\label{ref:RNDQXsIfZlLYA}(Jacquette, 2002, p.xiii; see e.g. Foschini, 2013).
%\parencites[][p.xiii]{jacquette_ontology_2002}[see, e.g.,][]{foschini_where_2013}.%
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Thus, ontology that represents the real world being always expressed in some form of language (which always is) is always incomplete with respect to the world, i.e., the world ``as is'' cannot be represented by something else, despite the close isomorphism between the reality and some logical aka ontological systems \footnote{All pure ontological systems are logical systems
%\label{ref:RNDQXsIfZlLYA}(Jacquette, 2002, p.xiii; see e.g. Foschini, 2013).
\parencites[][p.xiii]{jacquette_ontology_2002}[see e.g.,][]{foschini_where_2013}.%
} As a~result, there will always be some inconsistency between what exists and what a~given ontology represents, because ontology (even ontology \textbf{O}) will always be a~theory (of sorts), expressed in some specialized language (see ft. 8 ) about the world rather than the world itself. Thus, there will always be a~degree of incommensurability,\footnote{Incommensurability is not understood here in the same way as the incommensurability of paradigms or scientific theory in the works of Khun 
%\label{ref:RNDg7lp3HNe18}(1962)
\parencite*[][]{kuhn_structure_1962} %
 or Feyerabend 
%\label{ref:RNDdDGQ3TRwiM}(e.g. Ryan, 2002; Sankey, 1993; Oberheim and Hoyningen-Huene, 2018; Bird, 2000).
\parencites[e.g.,][]{ryan_feyerabend_2002}[for the incommensurability of paradigms, see also][]{sankey_kuhns_1993}[][]{oberheim_incommensurability_2018}[][]{bird_thomas_2000}. %
 Instead, incommensurability is understood here as a~general sense of not being entirely comparable according to some criteria.} or a~gap, between the ontology of synthetic systems and the reality of what exists. This gap may be narrowed but never entirely closed. Indeed, a~representation can never contain itself as a~part of what exists. We should keep this incommensurability in mind when building synthetic systems like~AI\textsubscript{E}.

\subsection*{The meta-ontology of AI}

Meta-ontology is a~relatively recently coined term with many interpretations. In philosophy meta-ontology denotes a~study of ontology, what it investigates, and what it is concerned with
%\label{ref:RNDOe0O2QLaGe}(e.g. Quine, 1960; Eklund, 2008; Berto and Plebani, 2015).
\parencites[e.g.,][]{quine_word_1960}[][]{eklund_picture_2008}[][]{berto_ontology_2015}. %
 Peter Van Inwagen 
%\label{ref:RNDlQ2vAbyD19}(1998),
\parencite*[][]{van_inwagen_meta-ontology_1998}, %
 the originator of the term, posited that the role of meta-ontology is to clarify the subject of ontology and explain how ontological claims can be interpreted. Francesco Berto and Matteo Plebani describe meta-ontology in terms of ``‘meta-X' as the inquiry on the central concepts and procedures of discipline X'' 
%\label{ref:RNDRhsH36U3p0}(Berto and Plebani, 2015, p.13),
\parencite[][p.13]{berto_ontology_2015}, %
 where ``X'' refers to ontology in this case. Meta-ontology asks what a~philosopher means when he asks ontological questions or questions about ontology 
%\label{ref:RND0XCU1LJYrl}(Eklund, 2008; Turner, 2014).
\parencites[][]{eklund_picture_2008}[][]{turner_metaontology_2014}. %
 Furthermore, meta-ontology is also concerned with ontological commitments and the truth conditions for a~given ontological theory 
%\label{ref:RNDPFqAz4nWnW}(Van Inwagen, 1998).
\parencite[][]{van_inwagen_meta-ontology_1998}. %
 From this perspective, meta-ontology would inquire as to what kind of ``things'' an ontological theory (i.e., ontology) is committed to.\footnote{We do not discuss Quine's meta-ontology as being specific to Quine's concept of ontology that is not considered here.}

Ontological commitment and truth condition are key terms for defining meta-ontology, and they are critical for differentiating between ontology and meta-ontology
%\label{ref:RNDIMrAPRSpV1}(Turner, 2014).
\parencite[][]{turner_metaontology_2014}. %
 On the conceptual level, the ontological commitment denotes what kind of ontology (i.e., what exists, what is) a~system or a~theory is supposed to represent. Gibson 
%\label{ref:RNDSpom5lKiCH}(2009, p.631)
\parencite*[][p.631]{gibson_ontological_2009} %
 states that ``theory is ontologically committed to an object only if that object occurs in all the ontologies of that theory.'' Thus, ontology may be committed to the existence of numbers, planets, subatomic particles, ghosts, values, ethics, and so on, anything as long as these ``things'' are recognized in ``all the ontologies of that theory'' 
%\label{ref:RND0EJlQmQ9Dk}(e.g. Eklund, 2008).
\parencite[e.g.,][]{eklund_picture_2008}. %
 While ontology is about existence, ``what is'', the ontological commitment is about representation of ``what is'' 
%\label{ref:RNDvi3QcENC1N}(Smith, 1998)
\parencite[][]{smith_origin_1998} %
 and the capacity to represent it.

Moreover, ontological commitment also includes verifying the criteria for this ontological commitment (i.e., verifying the truth of its existential claims). Ontological commitment can be reduced to the simple claim that ``A man is committed to the truth of whatever he asserts''
%\label{ref:RNDA5OgzdmbIa}(Searle, 1969, p.112).
\parencite[][p.112]{searle_speech_1969}.%
\footnote{See the critique of Searle by 
%\label{ref:RNDdyJS6eDQgJ}(Inwagen, 1991).
\parencite[][]{lepore_searle_1991}.%
} For some, however, particularly professional ontologists, Searle's take on ontological commitment seems perfunctory, but in the AI context, it provides a~simple (i.e., operational) means for judging the scope of an AI ontology. We do not dispute the robustness of Searle's claim on ontological commitment; here we take it as a~guide in practical applications.

Truth conditions are what makes an ontology correct, such that a~``theory is ontologically committed to an object only if that object occurs in all the ontologies of that theory''
%\label{ref:RND7ceF0NY8f3}(Gibson, 2009, p.631).
\parencite[][p.631]{gibson_ontological_2009}. %
 This statement was rephrased by Rayo 
%\label{ref:RND2olhyVRiGx}(2007)
\parencite*[][]{rayo_ontological_2007} %
 into the following claim: ``…for a~sentence to carry commitment to Fs is for the sentence's truth to demand of the world that it contain Fs.'' Gibson's claim therefore comes from a~philosophical perspective. It requires that a~theory (of ontology) can be expressed in sentences (of any language), so this truth condition boils down to an agreement between the theory and the world in question (the correspondence theory of truth\footnote{We are not going here into the discussion of the correspondence theory of truth or truth in general. We assume that for the systems (natural or artificial) ``being in the world'' (no Heideggerian connotations) there must be some relation, however tentative and limited holding between them and the real world they are immersed in, the relation of truth. This relation in a~case of evolution and synthetic systems is verified by systems' operational success 
%\label{ref:RND1clmQWWHtj}(see for example the discussion of truth relation in Bird, 2000).
\parencite[see for example the discussion of truth relation in][]{bird_thomas_2000}. %
 The correspondence theory of truth seems to trouble only philosophers of the anti-realist, skeptical persuasion, not computer scientists or evolutionary biologists 
%\label{ref:RNDL8KkhQlfH3}(Bird, 2000).
\parencite[][]{bird_thomas_2000}. %
 ``Operational success'' in applied sciences may be thought as playing the role of empirical proof in naturalized epistemology 
%\label{ref:RNDDKaQHD0jpK}(Bird, 2000, p.263).
\parencite[][p.263]{bird_thomas_2000}. %
 }), which may be the real world (as is the case of AI\textsubscript{E} systems) or an imaginary constructs or virtual realities whatever the domain of ontology is.

\subsection*{AI\textsubscript{E} Paradigm}

The term ``AI paradigm'' is used in many AI-related papers, discussions, and articles, with it carrying various meanings, so there is no agreement about what it should denote. In principle, authors simply define a~``paradigm'' in a~way that suits their narrative, thus confirming Ian Hacking's prediction that the term would become banal following Kuhn's publication
%\label{ref:RNDbr9uaPs2u1}(Hacking, 2012).
\parencite[][]{kuhn_introductory_2012}. %
 Thus, when we talk about AI paradigm we need to show precisely, to avoid misinterpretations, what we are talking about and how our definition of paradigm differs from, or is similar to, from other definitions.

Schopman
%\label{ref:RNDJc1x1QMFP2}(1986)
\parencite*[][]{schopman_artifical_1986} %
 suggests that AI has not developed a~specific paradigm, claiming that ``…no computational paradigm has yet been produced: there is no single generally accepted way to do AI'' 
%\label{ref:RNDPBgq3pJi3C}(Schopman, 1986, p.6).
\parencite[][p.6]{schopman_artifical_1986}. %
 Čaplinskas 
%\label{ref:RNDMRm8DpWtKy}(1998),
\parencite*[][]{caplinskas_ai_1998}, %
 however, defines three AI paradigms: the behaviorist paradigm, the agent paradigm, and the artificial life paradigm. Norvig 
%\label{ref:RNDlOHT6rHNA2}(1992; 2002),
\parencites*[][]{norvig_paradigms_1992}[][]{norvig_retrospective_2002}, %
 meanwhile, associates the term with Lisp programming to express the paradigm of Artificial Intelligence Programming as being equivalent to a~programming approach. Next, Cristianini (2014) distinguishes four AI paradigms: data-driven AI, statistical AI, knowledge-driven AI, and reasoning-and-search-based AI. Without much explanation as to why, Leary 
%\label{ref:RNDMgW31BP6Zx}(2017)
\parencite*[][]{leary_googles_2017} %
 claims that the new Google AI paradigm is machine learning, while in a~blog post titled ``The AI Paradigm Shift,'' Richardson 
%\label{ref:RNDt2zoIcErAM}(2018)
\parencite*[][]{richardson_ai_2018} %
 refers broadly to the AI paradigm as an approach to engineering AI systems, such as deep learning (DL), machine learning (ML), natural language processing (NLP), and robotics. 
%\label{ref:RNDcu5g1TESmD}(Hernández-Orallo et al., 2020, p.2522),
\parencite[][p.2522]{hernandez-orallo_ai_2020}, %
 meanwhile, claims that the concepts of AI paradigms have been used to denote ``broad families of technical or conceptual approaches: ‘symbolic' vs ‘connectionist', reasoning vs learning, expert systems vsagents.''

Much like Richardson, Romero
%\label{ref:RNDPl3RwoIA7v}(2021)
\parencite*[][]{romero_unpopular_2021} %
 refers to deep learning and machine learning as AI paradigms, while Yalçın 
%\label{ref:RNDmyfcw0kayw}(2021)
\parencite*[][]{osaba_artificial_2021} %
 refers to symbolic (i.e., a~human-readable, symbolic representation of problems, logic, and search) and sub-symbolic (i.e., an implicit representation derived from experience-based learning with no symbolic representation of rules and properties) representations as AI paradigms. Villar et al. 
%\label{ref:RNDY4KjMoWStM}(2021)
\parencite*[][]{xu_artificial_2021} %
 hint at relating the AI paradigm to versions of machine-learning and deep-learning methods. Meanwhile, Luhach Kumar and El\c{c}i Atilla 
%\label{ref:RND1BjB3dukU7}(2021)
\parencite*[][]{luhach_artificial_2021} %
 in their book use the term ``paradigm'' in several places but in different contexts, with its meaning being variably associated with programming, the applications of AI to smart computational cyberspaces, or execution paradigms associated with computer hardware. Next, Joseph Makokha 
%\label{ref:RNDS2MHw4L6Mk}(2021)
\parencite*[][]{makokha_artificial_2021} %
 distinguishes two AI paradigms, namely an AI-based one for rule-following methods and another one based on artificial neural network constructs. The term ``AI paradigm'' is often used to simply denote a~method for AI learning or knowledge acquisition, and this is how Yonguin Xu 
%\label{ref:RNDGRGBuNdPMm}(Xu et al., 2021)
\parencite[][]{xu_artificial_2021} %
 uses this term to denote~deep-learning approaches, such as supervised, unsupervised, and reinforced learning.

Several studies have posited that the current AI systems use two broad, conceptual constructs, namely the symbolic and the sub-symbolic
%\label{ref:RNDiWrk1ACHd1}(e.g. Searle, 1998; Harvey, 2013; Neapolitan and Jiang, 2018; Mitchell, 2019; Smith, 2019; Cole, 2020; Russell and Norvig, 2020; Wooldridge, 2021).
\parencites[e.g.,][]{searle_mind_1998}[][]{harvey_perspectives_2013}[][]{neapolitan_artificial_2018}[][]{mitchell_artificial_2019}[][]{smith_promise_2019}[][]{cole_chinese}[][]{russell_artificial_2020}[][]{wooldridge_road_2021}. %
 Thus, the symbolic paradigm reflects symbolic representations of a~priori defined concepts that may be implemented in various programming environments, while the sub-symbolic paradigm relates to less clearly defined concepts, such as the multi-dimensional probability weights on the connections within an artificial neural network. An approach under the sub-symbolic paradigm would therefore be implemented using one of the various machine learning (ML) technologies. These two paradigms can also be fused into a~neuro-symbolic paradigm 
%\label{ref:RNDTv0qtZz01g}(e.g. Bader and Hitzler, 2005; Garcez et al., 2019; Garcez and Lamb, 2020; Kautz, 2022)
\parencites[e.g.,][]{bader_dimensions_2005}[][]{garcez_neural-symbolic_2019}[][]{garcez_neurosymbolic_2020}[][]{kautz_third_2022} %
 that combines symbolic and sub-symbolic elements. As pointed out earlier, the term ``AI paradigm'' does not point to a~single software or hardware solution 
%\label{ref:RNDzI0NRHZm9F}(see Minsky, 1991; Russell and Norvig, 2020)
\parencites[see][]{minsky_logical_1991}[][]{russell_artificial_2020}%

In this paper, we follow the example of Searle, Harvey, Neapolitan and Jiang, Smith, Cole, Wooldrige, and Russell and Norwig in using the term ``AI paradigm'' to denote a~broad, conceptual construct that underlies AI systems. AI paradigm as is here defined does not imply a~specific implementation. The AI paradigm therefore allows for multiple implementations, formal structures, representations, programming methods, and processing algorithms,\footnote{We use the term algorithm in the general sense of a~procedure that can be conceptualized and implemented in a~computer. This use follows Knuth's definition of a~computational method as being ``A procedure that has all of the characteristics of an algorithm except that it possibly lacks finiteness may be called a~computational method''
%\label{ref:RNDXsmzRWvSXf}(Knuth, 2005, p.5).
\parencite[][p.5]{knuth_art_2005}.%
} with these all belonging to a~single paradigm.

\section*{The meta-ontology of AI\textsubscript{E} systems}
We have concluded that in philosophy, meta-ontology is the study of what ontology is all about. In other words, it is a~study of, or about, ontology, as well as what it investigates; after Berto and Plebani ``‘meta-X' as the inquiry on the central concepts and procedures of discipline X''where ``X'' refers to ontology in this case
%\label{ref:RNDc30wpAwAYr}(Berto and Plebani, 2015, p.13).
\parencite[][p.13]{berto_ontology_2015}. %
 We also concluded that ontological commitment and truth conditions represent the key concepts of meta-ontology, and these terms are critical for differentiating between ontology and meta-ontology.

The meta-ontology of AI\textsubscript{E} systems retains the core philosophical meaning (i.e., what ontology is committed to), but it diverges from the philosophical concept in the details, because it assumes the perspective of AI\textsubscript{E} systems. When we say it ``retains the core philosophical meaning,'' we mean that meta-ontology of AI\textsubscript{E} systems is concerned with ontology, or is about ontology (i.e., the philosophical meaning of meta-ontology). Contrary to the use in philosophy, however, the meta-ontology of the AI\textsubscript{E} focuses not on a~theory of ontology, as meta-ontologies in philosophy do, but rather on what the AI\textsubscript{E} represents (i.e., the real world).
